---
title: "Figure 3 - sQTL tally, correlation heatmap"
date: '2024-08-21'
code-fold: true
execute:
  include: false
  eval: false
  echo: false
  warning: false
  message: false
categories:
  - figures
  - manuscript
---

::: {.callout-note}

-  sQTL top SNPs enrichment in eQTL are plotted on [this page](../analysis/2024-08-05-GTEx-sqtl-enrich.qmd)

-  sQTL top SNPs eQTL beta vs. sQTL beta per tissue comparison are also on [this page](../analysis/2024-08-05-GTEx-sqtl-enrich.qmd)

- sQTL top SNPs eQTL beta vs. sQTL beta summarised heatmap is on [this page](../analysis/2024-07-25-plot-beta-beta-heatmap.html)
:::



```{r}
#| label: setup
#| eval: true
suppressMessages(library(tidyverse))
```



## sQTL tally (GTEx)

```{r}
#| include: true
#| eval: true
#| out-width: 100%
#| fig-cap: "GTEx sQTL tally"

knitr::include_graphics("../analysis/2024-06-10-gtex-sqtl-tally-by-tissue.svg")
```


## Heatmap of eQTL sQTL beta correlation


```{r}
#| include: true
#| eval: true
#| out-width: 100%
#| fig-cap: "Heatmap of eQTL sQTL beta correlation"
#| column: screen

knitr::include_graphics("../analysis/2024-07-25-heatmap-corr-beta-eqtl-sqtl.svg")
```

## sQTL pvalue enrichment in eQTL (selected tissues)


```{r}
#| include: true
#| eval: true
#| layout-ncol: 3
#| out-width: 100%

svg_files = dir('../smk-plots/gtex-sqtl-enrichment-v4', 
                pattern = 'qqplot\\.svg',
                full.names = TRUE)
svg_files = svg_files[str_detect(svg_files, "EBV|Artery.Tib|Testis")]

knitr::include_graphics(svg_files)

```


 
