---
title: "Check rerun of leafcutter2 on select GTEx tissues and Geuvadis"
date: 2024-9-2
format: html
categories: ['test', 'leafcutter2']
code-fold: true
execute:
  include: false
  eval: false
  cache: false
  echo: false
---


```{r}
#| label: functions
#| eval: true

suppressMessages(library(dplyr))
suppressMessages(library(tidyr))
suppressMessages(library(stringr))
suppressMessages(library(purrr))
suppressMessages(library(data.table))
suppressMessages(library(glue))

lf2 = function(f) {
    df = fread(f)
    dcols = colnames(df)[2:ncol(df)]
    m = df[, .(m = rowMeans(.SD)), .SDcols = dcols]
    df = separate(df[, .(chrom)], "chrom", c("chr", "start", "end", "clu", "itype"), sep = ":") %>% 
        mutate(strand = str_sub(clu, -1, -1)) %>% 
        mutate(clu = str_remove(clu, "_[\\+\\-]$")) %>%
        as.data.table
    df = cbind(df, m)
    return(df)
}

sqtl = function(folder, pattern) {
    files = dir(folder, pattern, full.names = TRUE)
    df = map_dfr(files, fread)
    df$itype = str_sub(df$phenotype_id, -2, -1)
    df = df[, .N, by = itype][, .(itype, N, P = round(N/sum(N), 3))]

    return(df)
}


```


# Introduction


A quick check of the leafcutter2 rerun on select GTEx tissues and Geuvadis. 

What's being compared: 

- leafcutter output
- mapped sqtls

GTEx: Kidney-Cortex, about 90 smaples. Geuvadis, EUR 300-400 samples.


# Leafcutter2 output


```{r}
#| eval: true
#| cache: true
lf.geu = lf2("../code/results/pheno/noisy/Geuvadis/EUR/leafcutter_perind_numers.counts.noise_by_intron.gz")

lf.geu.o = lf2("../code/results/pheno/noisy/Geuvadis/EUR-run-Jun-2024/leafcutter_perind_numers.counts.noise_by_intron.gz")

lf.gtx = lf2("../code/results/pheno/noisy/GTEx/Kidney-Cortex/leafcutter_perind_numers.counts.noise_by_intron.gz")
lf.gtx.o = lf2("../code/results/pheno/noisy/GTEx/Kidney-Cortex-run-may-2024/leafcutter_perind_numers.counts.noise_by_intron.gz")

```


## Geuvadis

```{r}
#| eval: true
#| include: true

lf.geu %>% head
lf.geu %>% dim
```

```{r}
#| eval: true
#| include: true

lf.geu.o %>% head
lf.geu.o %>% dim
```


Does the mean reads match for each intron?

```{r}
#| eval: true
#| include: true
full_join(
    lf.geu[, -c("clu")],
    lf.geu.o[, -c("clu")],
    by = c("chr", "start", "end", "strand"),
    suffix = c(".new", ".old")
) %>% 
    .[, .(.N), by = .(Match_m = m.new == m.old)]
```

Does the itypes match for each intron?

```{r}
#| eval: true
#| include: true
full_join(
    lf.geu[, -c("clu")],
    lf.geu.o[, -c("clu")],
    by = c("chr", "start", "end", "strand"),
    suffix = c(".new", ".old")
) %>% 
    .[, .(.N), by = .(Match_itype = itype.new == itype.old, itype.new, itype.old)] %>% 
    .[, .(Match_itype, itype.new, itype.old, N, P = round(N/sum(N), 3))] %>% 
    .[order(Match_itype, -P)]
```

## GTEx

```{r}
#| eval: true
#| include: true

lf.gtx %>% head
lf.gtx %>% dim
```

```{r}
#| eval: true
#| include: true

lf.gtx.o %>% head
lf.gtx.o %>% dim
```


Does the mean reads match for each intron?

```{r}
#| eval: true
#| include: true
full_join(
    lf.gtx[, -c("clu")],
    lf.gtx.o[, -c("clu")],
    by = c("chr", "start", "end", "strand"),
    suffix = c(".new", ".old")
) %>% 
    .[, .(.N), by = .(Match_m = m.new == m.old)]
```

Does the itypes match for each intron?

```{r}
#| eval: true
#| include: true
full_join(
    lf.gtx[, -c("clu")],
    lf.gtx.o[, -c("clu")],
    by = c("chr", "start", "end", "strand"),
    suffix = c(".new", ".old")
) %>% 
    .[, .(.N), by = .(Match_itype = itype.new == itype.old, itype.new, itype.old)] %>% 
    .[, .(Match_itype, itype.new, itype.old, N, P = round(N/sum(N), 3))] %>% 
    .[order(Match_itype, -P)]
```



# sQTLs

 
```{r}
#| eval: true
#| cache: true

sqtls.geu = sqtl("../code/results/qtl/noisy/Geuvadis/EUR/separateNoise/cis_100000/perm/", "*.addQval.txt.gz")
sqtls.geu.o = sqtl("../code/results/qtl/noisy/Geuvadis/EUR-run-may-2024/separateNoise/cis_100000/perm/", "*.addQval.txt.gz")

sqtls.gtx = sqtl("../code/results/qtl/noisy/GTEx/Kidney-Cortex/separateNoise/cis_100000/perm/", "*.addQval.txt.gz")
sqtls.gtx.o = sqtl("../code/results/qtl/noisy/GTEx/Kidney-Cortex-run-aug-2024/separateNoise/cis_100000/perm/", "*.addQval.txt.gz")
```

## Geuvadis

```{r}
#| eval: true
#| include: true
sqtls.geu[order(itype)]

sqtls.geu.o[order(itype)]
```


## GTEx

```{r}
#| eval: true
#| include: true
sqtls.gtx[order(itype)]

sqtls.gtx.o[order(itype)]
```




