---
title: "List of differentially spliced junctions"
date: 2024-9-5
format: html
code-fold: true
execute:
  include: false
  eval: false
  cache: false
  echo: false
---

>im making good progress on re-analyzing that comprative rna-seq data leafcutter2 style… but after discussing a little with yang today, we have an idea for a figure panel that would involve identifying a nice example of a conserved unproductive splice event that is also present in your GTEx cross-tissue differential splicing analysis. Can you share a list of junctions that are differentially spliced any pair of GTEx tissues that you looked at. Like, ideal would be a text file of human junction coordinates, and the tissue pairs from which the differential splicing was done, and a beta (so I can see if the direction of the effect is same in other other species).




::: {.callout-note}

## Notes


- Results here are before the recent implementation of foward splicing junction classifer and bug fixes.
:::


# List of differentially spliced junctions


List of differentialy spliced junctions are here: 


```{bash}
#| echo: false
#| eval: true
#| include: true

ls -1 /project/yangili1/cdai/SpliFi/data/ds_v_dge/*_v_*_data.rds | head
echo "Total number of pairs of comparisons"
ls -1 /project/yangili1/cdai/SpliFi/data/ds_v_dge/*_v_*_data.rds | wc -l

```




```{r}
#| label: setup
#| eval: true

suppressMessages(library(dplyr))
suppressMessages(library(tidyr))
suppressMessages(library(stringr))
suppressMessages(library(readr))
suppressMessages(library(purrr))
suppressMessages(library(data.table))
suppressMessages(library(glue))


FDR = 0.1
minPSI = 0.1

```


# Join differentially expressed genes and differentially spliced junctions

You can use this funciton to get the table of differentially spliced junctions and differentially expressed genes. 

```{r}
#| label: functions
#| eval: false
#| include: true
#| echo: true


get_DS_DGE_table <- function(rds, contrast, FDR = 1e-5, selection = 'all', plot=FALSE) {
    # rds: a single rds file, previousp rocessed in "../../SpliFi/prepdata.smk"
    # contrast: a string, the contrast name
    # return a joined table (on gene_names) of differential spliced junctions and diffenetially expressed genes
    data <- readRDS(rds)
    ds <- data$ds
    dge <- data$dge
    dge_dsc <- data$dge_dsc

    joined_dt <- inner_join(
        x = ds[abs(deltapsi) > .1 & `p.adjust` < FDR & ctype %in% c('PR', 'PR,UP'), 
        .(gene_id, gene_name, intron, cluster, itype, ctype, deltapsi)],
        y = dge[, .(gene_id, g_l2fc = log2FoldChange, g_z = stat, g_l2fcSE = lfcSE)],
        by = 'gene_id'
    )

    if (selection == 'maxClu') {
        joined_dt  <- joined_dt[
            , .(intron, deltapsi, g_l2fc, g_z, g_l2fcSE, rk = rank(-abs(deltapsi), ties.method = "random")), 
            by = .(gene_id, gene_name, cluster, itype, ctype)
            ][rk == 1, -c('rk')]
    } else if (selection == 'maxGene') {
        joined_dt <- joined_dt[
            , .(intron, deltapsi, g_l2fc, g_z, g_l2fcSE, rk = rank(-abs(deltapsi), ties.method = "random")), 
            by = .(gene_id, gene_name, itype)
            ][rk == 1, -c('rk')]
    }

    return(joined_dt)
}
```


# Other previously prepared data


## Clusters selected for plotting heatmaps

The Heatmap in Figure 2 is a common set of clusters across tissues. These clusters are selected because in at least some tissues (vs. others), they showed differential splicing and differential gene expression. 

Here's a set of chosen clusters. If a cluster has multiple UP introns, only the best one is selected (largest deltaPSI). Introns are also selected if deltaPSI and expression L2FC has different signs.


```{bash}
#| include: true
#| eval: true

ls -1 "/project/yangili1/cdai/SpliFi/data/ds_v_dge/chosen_clusters_dpsi_0_2.rds"

```

```{r}
#| include: true
#| eval: false

GetTopUpIntrons <- function(data, FDR_ds, FDR_dge, dPSI) {
  dge <- data$dge
  ds <- data$ds

  # if a cluster has multiple UP introns, only select the best 1
  ds <- ds[itype == 'UP' & ctype == 'PR,UP'][, rk := rank(-abs(deltapsi), ties.method = "first"), by = cluster][rk ==1][, rk := NULL][]
  ds <- ds[`p.adjust` < FDR_ds & abs(deltapsi) > dPSI,]

  dge <- dge[padj < FDR_dge,]
  
  ds_excl_cols <- c('itype', 'ctype', 'df', 'p', 'p.adjust','logef', 'loglr', 'status')
  dge_excl_cols <- c('baseMean', 'lfcSE', 'stat', 'pvalue', 'padj')
  chosen <- inner_join(
      x = ds[, -ds_excl_cols, with = FALSE],
      y = dge[, -dge_excl_cols, with = FALSE],
      by = "gene_id",
      suffix = c("_ds", "_dge")
    ) %>%
    .[deltapsi * log2FoldChange < 0, ]

  return(chosen)
}

```











































































