---
title: "Plot heatmap of effect size correlation between eQTL and sQTL (updated new vs. old pseudo count)"
date: 2024-8-11
format: html
categories: ['beta']
code-fold: true
execute:
  include: false
  eval: false
  cache: false
  echo: false
---


# Intro

The goal is to compute the correlations between sQTL beta and eQTL beta for top p vs. u sQTL top SNPs in a heatmap. It effectively summarises the 49 beta vs. beta scatter plots.

data:

1. p vs. u sQTL top SNPs
2. regression slope in sQTL mapping
3. regression slope in eQTL mapping

do this for each of the 49 tissues. Result is a 2x49 matrix where 

- row labels: u or p sQTL (2)
- col labels: tissue (49)
- values: correlation value of betas. 

plot heatmap: 

- split heatmap into 5 sub heatmaps, each is 2x10 (9) matrix




```{r}
#| label: libraries-funcs-setup
#| eval: true

suppressMessages(library(tidyverse))
suppressMessages(library(data.table))
suppressMessages(library(glue))
suppressMessages(library(ComplexHeatmap))

# functions
grab_corr_matrix <- function(rds_file) {
  data <- readRDS(rds_file)
  corrDF <- data$corrDF
  return(corrDF)
}

```


# Plots

## Correlation of effect sizes: eQTL and sQTL (updated pseudo count)

```{r}
#| label: files-params
#| eval: true


# rds files with correlation matrices (using new pseudo count method)
rds_files = list.files("/project/yangili1/cdai/splice-pub/smk-plots/gtex-sqtl-enrichment-v4-C", pattern = "*.rds", full.names = TRUE)
names(rds_files) <- rds_files %>%
  str_split("/") %>%
  map(pluck, 8) %>%
  unlist() %>%
  str_remove("\\.rds")


rds_files.old = list.files("/project/yangili1/cdai/splice-pub/smk-plots/gtex-sqtl-enrichment-v4", pattern = "*.rds", full.names = TRUE)
names(rds_files.old) <- rds_files.old %>%
  str_split("/") %>%
  map(pluck, 8) %>%
  unlist() %>%
  str_remove("\\.rds")


# gtex tissue names coded
gtex_tissue_code <- fread("/project/yangili1/cdai/splice-pub/analysis/gtex-tissue-code.csv", col.names = c("tissue", "code"), header = F)

```


```{r}
#| label: get-corr-matrix
#| eval: true
#| cache: true
corrDFs = map(rds_files, grab_corr_matrix)

df = corrDFs %>%
  map(select, estimate) %>%
  map(rownames_to_column, "type") %>%
  map(arrange, "type") %>%
  map(column_to_rownames, "type") %>%
  do.call(cbind, .)
colnames(df) <- names(corrDFs)

pval.df = corrDFs %>%
  map(select, pval) %>%
  map(rownames_to_column, "type") %>%
  map(arrange, "type") %>%
  map(column_to_rownames, "type") %>%
  do.call(cbind, .)
colnames(pval.df) <- names(corrDFs)

# order column based on UP correlation estimates
columnOrder <- df["UP", ] %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("tissue") %>%
  mutate(order = rank(UP, ties.method = "first")) %>%
  arrange(order) %>%
  pull(tissue)
```


```{r}
#| label: tbl-corr-heatmap-data-table
#| eval: true
#| include: true
#| tbl-cap: Correlation matrix of beta values between eQTL and sQTL

df[, columnOrder] %>%
  as.data.frame() %>%
  rownames_to_column("type") %>%
  gt::gt() %>%
  gt::fmt_number(decimals = 3)
```


```{r}
#| label: tbl-corr-heatmap-data-table-pval
#| eval: true
#| include: true
#| tbl-cap: Correlation pval matrix of beta values between eQTL and sQTL

# use gt to display p-values in scientific notation, cell fill color based on p-values
pval.df[, columnOrder] %>%
  as.data.frame() %>%
  rownames_to_column("type") %>%
  gt::gt() %>%
  gt::fmt_scientific(decimals = 2)

```



```{r}
# not used
# modified estimate based on p-value, used to plot heatmap differently for significant and non-significant correlations
df2 <- corrDFs %>%
  map(~ mutate(.x, estimate = if_else(pval < 0.01, estimate, 0))) %>%
  map(select, estimate) %>%
  map(rownames_to_column, "type") %>%
  map(arrange, "type") %>%
  map(column_to_rownames, "type") %>%
  do.call(cbind, .)
colnames(df2) <- names(rds_files)
```


```{r}
#| label: heatmap color scale

legendBreaks <- c(-.3, -.2, -.1, -0.0001, -0.00001, 0, 0.00001, 0.0001, 0.1)
colors <- c("#085193", "#4292c6", "#c6d8ef", "white", "white", "white", "white", "white", "#fc8d59")
colfunc <- circlize::colorRamp2(
  breaks = legendBreaks,
  colors = colors
)

```



```{r}
#| label: plot-heatmap

set.seed(123)
cluster_method <- "ward.D"
svg("2024-07-25-heatmap-corr-beta-eqtl-sqtl.svg", width = 20, height = 2)
df[, columnOrder] %>%
  as.matrix() %>%
  Heatmap(
    col = colfunc,
    show_parent_dend_line = FALSE,
    use_raster = TRUE,
    cluster_columns = FALSE, cluster_rows = F,
    column_split = c(rep("G1", 10), rep("G2", 10), rep("G3", 10), rep("G4", 10), rep("G5", 9)),
    column_labels = column_to_rownames(gtex_tissue_code, "tissue") %>% .[colnames(df[, columnOrder]),],
    height = unit(2, "cm"),
    heatmap_legend_param = list(title = "Correlation")
  )
dev.off()
```


```{r}
#| label: fig-corr-heatmap
#| column: screen
#| eval: true
#| include: true
#| fig-cap: Heatmap of correlation between eQTL and sQTL beta values (new pseudo count method)
#| out-width: 100%

knitr::include_graphics("2024-07-25-heatmap-corr-beta-eqtl-sqtl.svg")
```







## Correlation of effect sizes: eQTL and sQTL (old pseudo count)


```{r}
#| label: get-corr-matrix-old
#| eval: true
#| cache: true
corrDFs.old = map(rds_files.old, grab_corr_matrix)

df.old = corrDFs.old %>%
  map(select, estimate) %>%
  map(rownames_to_column, "type") %>%
  map(arrange, "type") %>%
  map(column_to_rownames, "type") %>%
  do.call(cbind, .)
colnames(df.old) <- names(corrDFs.old)

pval.df.old = corrDFs.old %>%
  map(select, pval) %>%
  map(rownames_to_column, "type") %>%
  map(arrange, "type") %>%
  map(column_to_rownames, "type") %>%
  do.call(cbind, .)
colnames(pval.df.old) <- names(corrDFs.old)

# order column based on UP correlation estimates
columnOrder <- df.old["UP", ] %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("tissue") %>%
  mutate(order = rank(UP, ties.method = "first")) %>%
  arrange(order) %>%
  pull(tissue)
```


```{r}
#| label: tbl-corr-heatmap-data-table-old
#| eval: true
#| include: true
#| tbl-cap: Correlation matrix of beta values between eQTL and sQTL (old pseudo count)

df.old[, columnOrder] %>%
  as.data.frame() %>%
  rownames_to_column("type") %>%
  gt::gt() %>%
  gt::fmt_number(decimals = 3)
```


```{r}
#| label: tbl-corr-heatmap-data-table-pval-old
#| eval: true
#| include: true
#| tbl-cap: Correlation pval matrix of beta values between eQTL and sQTL (old pseudo count)

# use gt to display p-values in scientific notation, cell fill color based on p-values
pval.df.old[, columnOrder] %>%
  as.data.frame() %>%
  rownames_to_column("type") %>%
  gt::gt() %>%
  gt::fmt_scientific(decimals = 2)

```



```{r}
#| label: plot-heatmap-old

set.seed(123)
cluster_method <- "ward.D"
svg("2024-07-25-heatmap-corr-beta-eqtl-sqtl-old.svg", width = 20, height = 2)
df.old[, columnOrder] %>%
  as.matrix() %>%
  Heatmap(
    col = colfunc,
    show_parent_dend_line = FALSE,
    use_raster = TRUE,
    cluster_columns = FALSE, cluster_rows = F,
    column_split = c(rep("G1", 10), rep("G2", 10), rep("G3", 10), rep("G4", 10), rep("G5", 9)),
    column_labels = column_to_rownames(gtex_tissue_code, "tissue") %>% .[colnames(df.old[, columnOrder]),],
    height = unit(2, "cm"),
    heatmap_legend_param = list(title = "Correlation")
  )
dev.off()
```



```{r}
#| label: fig-corr-heatmap-old
#| column: screen
#| eval: true
#| include: true
#| fig-cap: Heatmap of correlation between eQTL and sQTL beta values (old pseudo count method)
#| out-width: 100%

knitr::include_graphics("2024-07-25-heatmap-corr-beta-eqtl-sqtl-old.svg")
```
































